---
title: "Your Name"
---

# Poisson Processes with Outliers

## Sampling Distribution of the Sample Mean

Answer goes here

## Power of the Test

Answer goes here

## Posterior Distribution

Answer goes here

## Summary

Answer goes here

# Current Population Survey

## Frequentist Perspective

Answer goes here

## Getting CPS Data

This is a chunk of R code that gets executed when you knit the RMarkdown file into a HTML or PDF file.
You can insert more of them using the drop-down menu above and to the right.
```{r}
library(haven)
CPS <- as_factor(read_dta(dir(pattern = "^cepr_.*dta$"))) 
# as_factor changes the categorical variables in Stata to R factors
```

## Selection

This R chunk evaluates but suppresses the R code in it and gets rid of the formatting so you just see
plain Stan code in the knitted HTML or PDF file
```{r, echo = FALSE, comment = ""}
writeLines(readLines(dir(pattern = "stan$"))) # works if you save your Stan file in the same directory
```

## Prior Predictive Checking

Answer goes here

## Estimating a Probit Model

Answer goes here

## Inference

Answer goes here
